---
title: 'Math 170S Homework #7'
author: "Damien Ha"
date: "2023-03-04"
output:
  pdf_document:
    toc: true
    toc_depth: 3
---

# Problem 1

## Part A

```{r}
sample1 <- c(20, 25, 18, 27, 32, 39, 21, 26, 19, 28, 33, 38, 22, 27, 23, 35, 31)
sample2 <- c(39, 21, 23, 24, 29, 28, 27, 29, 31, 33, 36, 37)

t.test(sample1, sample2, var.equal = FALSE, conf.level = 0.95)
```

At a level of $\alpha = 0.05$, the p-value is 0.2917 which is greater than $\alpha = 0.05$, so we fail to reject the null hypothesis. There is not enough evidence to conclude that the two samples are not equal.


## Part B

```{r}
sample1 <- c(25, 31, 22.5, 33.5, 40, 48.7, 26.25, 32.5, 23.5, 35, 41.25, 47.5, 
             27.5, 33.7, 28.75, 43.7, 39, 48, 54, 56)
sample2 <- c(48, 26, 28.7, 30, 36.2, 35, 33.7, 36.2, 38.7, 41.5, 45, 46.5, 25, 26, 27)

t.test(sample1, sample2, var.equal = FALSE, conf.level = 0.95)
```

At a level of $\alpha = 0.05$, the p-value is 0.523 which is greater than $\alpha = 0.05$, so we fail to reject the null hypothesis. There is not enough evidence to conclude that the two samples are not equal.


## Part C

```{r}
sample1 <- c(209, 277, 221, 215, 247.7, 253.7, 218, 212, 244.7, 250.7, 215.0, 209)
sample2 <- c(265, 201, 228, 289, 277, 289, 288, 278, 277, 265, 264, 266, 298, 291, 270)

t.test(sample1, sample2, var.equal = FALSE, conf.level = 0.95)
```

Our p-value is 0.0003175. At a level of $\alpha = 0.05$, this is significant. In addition, our confidence interval does not include 0, suggesting that the true difference between the samples is not 0. Because of this, we reject the null hypothesis and conclude that the samples are not equal.



# Problem 2

## Part A
```{r}
sample1 <- c(55, 56, 59, 51, 82, 88, 72, 84, 71, 80, 66, 67, 52, 81, 57, 72, 69)
sample2 <- c(44, 43, 48, 56, 54, 53, 52, 41, 40, 43, 43, 44, 45, 45, 52, 51, 50)

t.test(sample1, sample2, paired = TRUE, alternative = "greater", conf.level = 0.99)
```

The p-value is 0.000002713 which is less than $\alpha = 0.01$, so we reject the null hypothesis. The mean of sample 1 is greater than that of sample 2 at the 99% confidence level. 

## Part B
```{r}
sample1 <- c(140, 142, 150, 130.1, 209, 224, 183, 214, 181, 204, 168, 170, 133, 
             206, 145, 183, 179)
sample2 <- c(105, 102, 115, 135.8, 130, 128, 125, 97, 95, 102, 102, 105, 108, 
             107, 125, 123, 120)


t.test(sample1, sample2, paired = TRUE, alternative = "greater", conf.level = 0.99)
```

The p-value is 0.000000578 which is less than $\alpha = 0.01$, so we reject the null hypothesis.  The mean of sample 1 is greater than that of sample 2 at the 99% confidence level. 

## Part C

```{r}
sample1 <- c(53.4, 52.88, 58.13, 62.27, 70.9, 72.5, 65.7, 60.7, 55, 60, 56, 57, 53, 62)
sample2 <- c(61, 63, 71, 55, 53, 59, 63, 61, 55, 58, 63, 58, 57, 56)

t.test(sample1, sample2, paired = TRUE, alternative = "greater", conf.level = 0.99)
```

The p-value is 0.4229 which is greater than $\alpha = 0.01$, so we fail to reject the null hypothesis. There is not enough evidence to suggest te mean of sample 1 is greater than sample 2 at a 99% considence level.



# Problem 3

## Part A

```{r}
sample1 <- c(55, 56, 59, 51, 82, 88, 72, 84, 71, 80, 66, 67, 52, 81, 57, 72, 69)
sample2 <- c(44, 43, 48, 56, 54, 53, 52, 41, 40, 43, 43, 44, 45, 45, 52, 51, 50)

wilcox.test(sample1, sample2, alternative = "greater", conf.level = 0.99)
```

The p-value is 0.000002264, which is less than $\alpha = 0.01$, so we reject the null hypothesis.  The mean of sample 1 is greater than that of sample 2 at the 99% confidence level. 

## Part B

```{r}
sample1 <- c(140, 142, 150, 130.1, 209, 224, 183, 214, 181, 204, 168, 170, 133, 
             206, 145, 183, 179)
sample2 <- c(105, 102, 115, 135.8, 130, 128, 125, 97, 95, 102, 102, 105, 108, 
             107, 125, 123, 120)

wilcox.test(sample1, sample2, alternative = "greater", conf.level = 0.99)
```

The p-value is 0.000000495, which is less than $\alpha = 0.01$, so we reject the null hypothesis.  The mean of sample 1 is greater than that of sample 2 at the 99% confidence level. 

## Part C

```{r}
sample1 <- c(53.4, 52.88, 58.13, 62.27, 70.9, 72.5, 65.7, 60.7, 55, 60, 56, 57, 53, 62)
sample2 <- c(61, 63, 71, 55, 53, 59, 63, 61, 55, 58, 63, 58, 57, 56)

wilcox.test(sample1, sample2, alternative = "greater", conf.level = 0.99)
```

The p-value is 0.573 which is greater than $\alpha = 0.01$, so we fail to reject the null hypothesis. There is not enough evidence to suggest te mean of sample 1 is greater than sample 2 at a 99% considence level.



# Problem 4

## Part A

```{r}
table <- matrix(c(34, 47, 55, 49, 39, 41, 52, 42, 32, 44, 48, 45), 
                nrow = 3, ncol = 4, byrow = TRUE)
rownames(table) <- c("M1", "M2", "M3")
colnames(table) <- c("A", "B", "C", "D")

chisq.test(table)
```

The p-value is 0.9647 which is greater than $\alpha = 0.05$, so we fail to reject the null hypothesis. There isn't enough evidence to suggest that there is association between the variables.

## Part B

```{r}
table <- matrix(c(123, 156, 228, 109, 144, 201, 111, 148, 212), 
                nrow = 3, ncol = 3, byrow = TRUE)
rownames(table) <- c("M1", "M2", "M3")
colnames(table) <- c("A", "B", "C")

chisq.test(table)
```

The p-value is 0.997 which is greater than $\alpha = 0.05$, so we fail to reject the null hypothesis. There isn't enough evidence to suggest that there is association between the variables.

## Part C

```{r}
table <- matrix(c(405, 656, 556, 655, 310, 608, 534, 387, 409, 626, 534, 456, 
                  400, 672, 443, 542), 
                nrow = 4, ncol = 4, byrow = TRUE)
rownames(table) <- c("M1", "M2", "M3", "M4")
colnames(table) <- c("A", "B", "C", "D")

chisq.test(table)
```

The p-value is 0.00000000001991 which is less than $\alpha = 0.05$, so we reject the null hypothesis. At a 95% confidence interval, there is evidence to suggest that the variables are associated.




